\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-03 papers/jfp/5-rebuilders.tex
Minor rewording

An alternative way to determine if a key is dirty is to record the
values/hashes of dependencies used last time, and if something has changed, the
key is dirty and must therefore be rebuilt~--~in essence, keeping a \emph{trace}
which we can use to \emph{verify} existing values. Taking the example from
Fig.~\ref{fig-bazel}(c), we might record that the key \cmd{util.o} (at hash 8)
depended on the keys \cmd{util.c} (at hash 7) and \cmd{util.h} (at hash 2). Next
time round, if the scheduler decides that it is time for \cmd{util.o} to be
rebuilt and all keys still have the same hashes as in the recorded trace, there
is nothing to do, and we can skip rebuilding. If any of the hashes is different,
we rebuild \cmd{util.o}, and record a trace with the new values.

For traces, there are two essential operations~--~adding a new trace to the
trace store, and using the traces to determine if a key needs rebuilding.
Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} can store
only hashes, of type \hs{Hash}~\hs{v}, of those values. Since the verifying
trace persists from one build to the next -- it constitutes the build system's
``memory'' -- it is helpful for it to be of modest size. After successfully
building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
passing the key, the hash of its value, and the list of hashes and dependencies.

@@

\begin{itemize}
\item Find the latest (with respect to the field \hs{built}) step trace matching
  \hs{k}. If it does not exist, return \hs{False}: \hs{k} was never built
  before and cannot be verified.
\item If \hs{h} does not equal the \hs{result} field of the trace, return
  \hs{False}: the current \hs{k}'s value was written externally and needs
  rebuilding.
\item For each key \hs{d} in \hs{depends}:\\
  $\circ$ Make sure \hs{d} is up-to-date, suspending the current task if needed;\\
  $\circ$ If \hs{d}'s latest \hs{changed} time is greater than \hs{k}'s
    \hs{built} time, return \hs{False}.
\item Return \hs{True}: the current \hs{k}'s value is up-to-date.
\end{itemize}

@@

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we require the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

@@

In practice, many cloud build systems store hashes of values in the trace store,
i.e. use \hs{Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v)} trace entries just like the
verifying traces, and have a separate content-addressable cache which associates
hashes with their actual contents.

@@

\begin{itemize}
  \item \textbf{Tasks must be deterministic}: If the tasks are not
        \emph{deterministic} then it is possible to violate correctness, as
        illustrated by an example in~\S\ref{sec-cloud-aspects}, see
        Fig.~\ref{fig-frankenbuild}.
  \item \textbf{No early cutoff}: Deep constructive traces cannot support early
        cutoff (\S\ref{sec-background-shake}), since the results of intermediate
        computations are not considered.
\end{itemize}

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
cost of losing such simple shallow builds and still requiring deterministic tasks.

>>>>

An alternative way to determine if a key is dirty is to record the
values/hashes of dependencies used last time, and if something has changed, the
key is dirty and must be rebuilt~--~in essence, keeping a \emph{trace}
which we can use to \emph{verify} existing values. Taking the example from
Fig.~\ref{fig-bazel}(c), we might record that the key \cmd{util.o} (at hash 8)
depended on the keys \cmd{util.c} (at hash 7) and \cmd{util.h} (at hash 2). Next
time round, if the scheduler decides that it is time for \cmd{util.o} to be
rebuilt and all keys still have the same hashes as in the recorded trace, there
is nothing to do, and we can skip rebuilding. If any of the hashes is different,
we rebuild \cmd{util.o}, and record a trace with the new values.

For traces, there are two essential operations~--~adding a new trace to the
trace store, and using the trace store to determine if a key needs rebuilding.
Assuming a store of verifying traces \hs{VT}~\hs{k}~\hs{v}, the operations are:

@@

\noindent
Rather than storing (large) values \hs{v}, the verifying trace \hs{VT} can store
only hashes of those values, with type \hs{Hash}~\hs{v}. Since the verifying
trace persists from one build to the next -- it constitutes the build system's
``memory'' -- it is helpful for it to be of modest size. After successfully
building a key, we call \hs{recordVT} to add a record to the current \hs{VT},
passing the key, the hash of its value, and the list of hashes and dependencies.

@@

\begin{itemize}
\item Find the latest (with respect to the field \hs{built}) step trace matching
  \hs{k}. If it does not exist, return \hs{False}: \hs{k} was never built
  before and cannot be verified.
\item If \hs{h} does not equal the \hs{result} field of the trace, return
  \hs{False}: the current \hs{k}'s value was changed externally and thus needs
  rebuilding.
\item For each key \hs{d} in \hs{depends}:\\
  $\circ$ Make sure \hs{d} is up-to-date, suspending the current task if needed;\\
  $\circ$ If \hs{d}'s latest \hs{changed} time is greater than \hs{k}'s
    \hs{built} time, return \hs{False}.
\item Return \hs{True}: the current \hs{k}'s value is up-to-date.
\end{itemize}

@@

\noindent
The function \hs{recordCT} looks like \hs{recordVT}, but instead of just passing
the hash of the resulting value, we pass the actual value. The \hs{verifyVT}
has been replaced with \hs{constructCT}, which instead of taking the hash of the
current value as \emph{input}, returns a list of possible values as
\emph{output} -- there may be more than one, because some build tools are
non-deterministic, see~\S\ref{sec-non-determinism}.

@@

In practice, many cloud build systems store hashes of values in the trace store,
i.e. use \hs{Trace}~\hs{k}~\hs{v}~\hs{(Hash}~\hs{v)} entries just like
verifying traces, and have a separate content-addressable cache which associates
hashes with their actual contents.

@@

\begin{itemize}
  \item \textbf{Tasks must be deterministic}: If the tasks are not
        \emph{deterministic} then it is possible to violate correctness, as
        illustrated by the example in~\S\ref{sec-cloud-aspects}, see
        Fig.~\ref{fig-frankenbuild}.
  \item \textbf{No early cutoff}: Deep constructive traces cannot support early
        cutoff (\S\ref{sec-background-shake}), since the results of intermediate
        computations are not considered.
\end{itemize}

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$. By picking values of $n$ in between we would regain some early cutoff, at the
cost of losing such simple shallow builds, while still requiring deterministic tasks.
